Overlay databank unlocks data-driven analyses of biomolecules for all

### 2023-10-09 main.tex
Updates from Overleaf

\begin{abstract}
Tools based on artificial intelligence (AI) are currently revolutionising many fields, yet their applications are often limited by the lack of suitable training data in programmatically accessible format. Here we propose an effective solution to make data scattered in various locations and formats accessible for data-driven and machine learning applications using the overlay databank format. To demonstrate the practical relevance of such approach, we present the NMRlipids Databank---a community-driven, open-for-all database featuring programmatic access to quality-evaluated atom-resolution molecular dynamics simulations of cellular membranes. Cellular membrane lipid composition is implicated in diseases and controls major biological functions, but membranes are difficult to study experimentally due to their intrinsic disorder and complex phase behaviour. While MD simulations have been useful in understanding membrane systems, they require significant computational resources and often suffer from inaccuracies in model parameters. Here, we demonstrate how programmable interface for flexible implementation of data-driven and machine learning applications, and rapid access to simulation data through a graphical user interface, unlock possibilities beyond current MD simulation and experimental studies to understand cellular membranes. The proposed overlay databank concept can be further applied to other biomolecules, as well as in other fields where similar barriers hinder the AI revolution.

%Cellular membrane lipid composition is implicated in diseases and controls major biological functions, but membranes are difficult to study experimentally due to their intrinsic disorder and complex phase behaviour. Molecular dynamics (MD) simulations have been useful in understanding membrane systems, but they require significant computational resources and often suffer from inaccuracies in model parameters. Applications of data-driven and machine learning methods, currently revolutionising many fields, remain of limited use for membrane systems due to the lack of suitable training sets. Here we present the NMRlipids Databank---a community-driven, open-for-all database featuring programmatic access to quality-evaluated atom-resolution MD simulations of lipid bilayers. The NMRlipids Databank will benefit scientists in different disciplines by providing automatic ranking of simulations based on their quality against experiments, programmable interface for flexible implementation of data-driven and machine learning applications, and rapid access to simulation data through a graphical user interface. To demonstrate how it unlocks possibilities beyond current MD simulation studies, we analyzed the NMRlipids Databank to reveal how anisotropic diffusion of water and cholesterol flip-flop rates depend on membrane properties.

\end{abstract}
\begin{document}

\flushbottom
\maketitle
% * <john.hammersley@gmail.com> 2015-02-09T12:07:31.197Z:
%
%  Click the title above to edit the author information and abstract
%
\thispagestyle{empty}

@@

Cellular membranes contain hundreds of different types of lipid molecules that regulate the membrane properties, morphology, and biological functions~\cite{vanmeer08,Lorent:2020a,Slatter:2016a}. Membrane lipid composition is implicated in diseases, such as cancer and neurodegenerative disorders, and therapeutics that affect membrane compositions are emerging~\cite{torres21}. However, biomembranes are often difficult to study experimentally, because they are complex mixtures of proteins and lipids in disordered fluid state with complicated phase behaviour at biological conditions. Molecular dynamics (MD) simulations can be used to model biomembranes in detail, but the computational cost for simulating all possible biological membrane compositions would be formidable. For those reasons, data-driven and machine-learning-based models that predict biomembrane properties will benefit wide range of fields covering academia and industry, from cell membrane biology to lipid nanoparticle formulations.
%For those reasons, the detailed connections between complex lipid interactions and biological functions taking place in or around membranes remain poorly understood. 
%Molecular dynamics (MD) simulations have been particularly useful in understanding membrane systems, although their accuracy has often been compromised by artefacts such as the quality of model parameters~\cite{antila22b,gupta22}. Presently, the accuracy of models is becoming increasingly important as researches are progressing from simulations of individual molecules to simulating whole organelles or even cells using interdisciplinary approaches~\cite{johnson15,thornburg22,gupta22}. Such systems exhibit intricate emergent behaviour making inaccuracies more difficult to detect, and accumulation of even modest errors may have a dramatic impact on the conclusions drawn.

%In contrast to experimental structural biology, where standard protocols to share and quality-evaluate resolved biomolecular structures are firmly established by the Protein Data Bank (PDB)~\cite{montelione13}, equivalent best practices are yet to be defined for MD simulations. The importance of such approaches is widely recognized~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20} and data-sharing solutions are emerging for proteins in solution~\cite{meyer10,kamp10}, proteins in membranes~\cite{newport19,espigares20,leston22}, nucleic acids~\cite{hospital16}, nucleic acids and proteins~\cite{bekker20}, cyclodextrins~\cite{mixcoha16}, COVID-19--involved macromolecules~(\href{https://bioexcel-cv19.bsc.es}{bioexcel-cv19.bsc.es}). However, automatic quality evaluation of simulation data~\cite{meyer10,hospital16} and programmatic access are still rare. In particular, tools for automatic quality evaluation of membrane simulations, or training sets for machine learning models of membrane-containing systems, are not yet available. Recent advances using machine learning approaches utilizing publicly available databanks, for example to solve protein structures~\cite{jumper21}, emphasize the increasing importance of such resources.

Here we present the NMRlipids Databank --- a community-driven, open-for-all database featuring programmatic access to atom-resolution MD simulations of lipid bilayers. To demonstrate its advantages over existing approaches, we show how information on rare phenomena that are beyond the scope of standard MD simulation investigations can be gleaned from the Databank. In addition, we demonstrate the immediate relevance of NMRlipids Databank in extending the scope of MD simulations to new fields: Using a data-driven approach, we are able to analyse how anisotropic diffusion of water depends on membrane properties; this benefits understanding in magnetic resonance imaging (MRI)~\cite{topgaard20} and pharmacokinetics~\cite{nitsche19}, where MD simulations have, until now, been rarely applied.
%Wider adaptation of the NMRlipids Databank will open even more possibilities. 
Furthermore, the Databank performs automatic quality evaluation of membrane simulations, which facilitates the selection of best-performing models for each given application and accelerates the development of simulation parameters and methodology. Notably, the overlay-databank and open-collaboration approaches would facilitate the collecting of community-contributed data and providing programmatic access to them also in fields other than membrane simulations. For example, substantial biomolecular MD simulation data are already available but in scattered locations and formats~\cite{tiemann23}. The technical advances presented here will immediately benefit making these data programmatically accessible. Furthermore, the overlay databank configuration will benefit a broad range of fields where access to training data is a bottleneck for building AI-based tools.
%NMRlipids databanksolutions presented as well as broader communities that 
%enables users to apply their data-driven approaches on the Databank, thus facilitating the creation of new tools for (and by) researchers in a .
% to computational chemistry and machine learning.

%While the NMRlipids Databank currently contains only lipid bilayer systems, its key elements that enable programmatic access to large-scale MD simulation data can be applied also for other molecules, such as disordered proteins or sugars.
%%The powerful combination of an overlay databank structure and a community-wide open collaboration can be used to build databases that enable data-driven and machine learning applications also in other fields, where storage of raw data requires significant resources, best practices in the field are not yet defined, and incentives to share data do not exist---such as assignment of NMR spectra~\cite{klukowski22}. 
%The powerful combination of an overlay databank structure and the community-wide open collaboration  described here is potentially useful for building databases that enable data-driven and machine learning applications also in other fields. This is especially true for research activities where storage of raw data requires significant resources, best practices in the field are not yet defined, and incentives to share data do not exist---such as assignment of NMR spectra~\cite{klukowski22}.

@@

The overlay structure of the NMRlipids Databank, illustrated in Fig.~\ref{fig:overlay}A, is designed to enable efficient upcycling of MD simulations for data-driven and machine learning applications with minimal investment on new infrastructure. Raw simulation data in the {\it Data layer} can be stored in any publicly available location with long term stability
%figshare (\href{https://figshare.com}{figshare.com}), or NIRD research data archive (\href{https://archive.norstore.no}{archive.norstore.no})---
and with permanent links to the data, such as digital object identifiers (DOIs), such as Zenodo (\href{https://zenodo.org}{zenodo.org}). 
%or uniform resource names (URNs). 
The {\it Databank layer} (\href{https://github.com/NMRlipids/Databank}{github.com/NMRlipids/Databank}) is the core of the Databank containing all the relevant information about the simulations: links to the raw data, relevant metadata describing the systems, universal naming conventions for lipids and their atoms, quality evaluation of simulations against experimental data, and the computer programs to create the entries and to analyse the five basic properties extracted from all simulations (area per lipid, C--H bond order parameters, X-ray scattering form factors, membrane thickness, and equilibration times of principal components). Also the values for these five basic properties are stored in the {\it Databank layer}.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\linewidth]{Figures/overlay3.pdf}
    \caption{Overview of the NMRlipids Databank. 
    \textbf{A} Schematic presentation of the overlay structure used in the NMRlipids Databank. A more detailed structure of the \textit{Databank layer} is shown in Fig.~\ref{DatabankStructure} in the SI.
    \textbf{B} Distribution of lipids present in the trajectories of the Databank. `Others' lists lipids occurring in five or fewer simulations. 
    \textbf{C} Distribution of force fields in the simulations in the Databank. References for each force field are given in Table~\ref{tab:ForceFields}.
    \textbf{D} Flowchart for performing an analysis of properties through all MD simulations in the NMRlipids Databank using the API.
    \textbf{E} Flowchart for accessing results calculated from the NMRlipids Databank and stored to the \textit{Application layer}.
    \textbf{F} Currently available single lipid component and binary mixtures in the NMRlipids Databank. Colorbar shows the number of available simulations with the darkest green indicating three or more.  
    }
    \label{fig:overlay}
\end{figure}

@@

\subsection{NMRlipids Databank-API: programmatic access to the MD simulation data}\label{section:access}
The NMRlipids Databank-API, available at \href{https://github.com/NMRlipids/Databank}{github.com/NMRlipids/Databank}, provides programmatic access to all simulation data in the NMRlipids Databank through application programming interface (API). This enables wide range of novel data-driven applications---from construction of machine learning models that predict membrane properties, to automatic analysis of virtually any property across all simulations in the Databank. The flowchart in Fig.~\ref{fig:overlay}D illustrates the practical implementation of such an analysis. After cloning the databank repository to a local computer, raw data of each simulation can be accessed by iterating through the README.yaml files in the {\it Databank layer} that contain links pointing to the locations of simulation trajectories and other relevant files. Simulations can then be automatically analyzed with the help of the README.yaml and the mapping files that associate the specific naming conventions in each simulation with the universal molecule and atom names used by the Databank. Finally, the analysis results are stored using the same structure as in the {\it Databank layer}. Practical examples of codes that utilize the NMRlipids Databank-API, as well as a template for a new analysis, can be found from locations listed in Table~\ref{tab:codes}.

@@

\subsection{Applications of the NMRlipids Databank}
\subsubsection{Selecting simulation parameters using NMRlipids Databank: Best models for most abundant neutral membrane lipids}
%Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19,??}, but the lack of universal quality measure for membrane simulations complicates the estimations of reliability of simulations. 
%
%For example, OPLS3e parameters overcome CHARMM36 in structural quality for POPC but predicts overestimated ion binding~\cite{kurki22}. On the other hand, CHARMM36 gives the best description for lipid headgroup conformational ensembles~\cite{bacle21}, while GROMOS-CKP parameters best capture the membrane packing in POPS bilayers~\cite{antila22b}. 
%
Molecular dynamics (MD) simulations have been particularly useful in understanding membrane systems, although their accuracy has often been compromised by artefacts such as the quality of model parameters~\cite{antila22b,gupta22}. Presently, the accuracy of models is becoming increasingly important as researches are progressing from simulations of individual molecules to simulating whole organelles or even cells using interdisciplinary approaches~\cite{johnson15,thornburg22,gupta22}. Such systems exhibit intricate emergent behaviour making inaccuracies more difficult to detect, and accumulation of even modest errors may have a dramatic impact on the conclusions drawn.
To minimize the detrimental consequences of artificial MD simulation results for their applications, the quality of lipid bilayer MD simulations has to be carefully assessed~\cite{antila22b}. This can be done, for example, against the C--H bond order parameters from NMR spectroscopy~\cite{bacle21,Wur23} and the form factors from X-ray scattering~\cite{ollila16}, although it requires comparisons between large number of simulations, which is laborious even with collaborative approaches~\cite{botan15,catte16,antila19,bacle21}. To streamline this process, we have defined quantitative quality measures that enable automatic ranking of lipid bilayer simulations based on their quality against experiments. Conformational ensembles of individual lipid molecules are evaluated in the NMRlipids Databank by first calculating the probabilities for each C--H bond order parameter to locate within experimental error and then averaging the possibilities over different lipid segments ($P^{\mathrm{hg}}$, $P^{\mathrm{sn1}}$,$P^{\mathrm{sn2}}$, and $P^{\mathrm{total}}$). Furthermore, the quality against X-ray scattering experiments ($FF_q$) is estimated as the difference in the experimental and simulated locations of the first form factor minimum. These measures can be used to evaluate membrane dimensions, as the form factor minima locations and acyl chain C--H bond order parameters correlate with the membrane lateral packing and thickness (Figs.~\ref{fig:quality}G,~\ref{fig:QualityCorrelationsSI}, and~\ref{fig:sizedependence}). In addition, the ergodicity of conformational sampling of lipids is estimated by calculating $\tau_\mathrm{rel}$, the convergence time of the slowest principal component divided by the simulation length. Here we demonstrate how the automatic simulation-quality evaluation and the NMRlipids Databank-API enable rapid selection of the best models for simulations of membranes with the two most biologically-abundant neutral membrane lipids~\cite{vanmeer08}: POPC and POPE.

As Fig.~\ref{fig:quality}A illustrates, predictions for the lateral packing of POPC and POPE membranes in terms of area per lipid, one of the most important parameters used to characterize cellular membranes, diverge between different force fields. To select the most realistic currently available force fields to describe POPC--POPE membranes, we first order simulations in the Databank according to their C--H bond order parameter quality ranking (Fig.~\ref{fig:top50simulations}). From these results, we then pick the force fields occurring in Fig.~\ref{fig:quality}A (that is: force fields for which both POPC and POPE simulations exist in the Databank) and rank them according to the quality of the \textit{sn}-1 chain of POPC (Fig.~\ref{fig:quality}B) and of POPE (Fig.~\ref{fig:quality}C). Simulations with $\tau_\mathrm{rel}$ clearly above one (larger than 1.3) are discarded. Because the membrane packing correlates with the average order parameter of the \textit{sn}-1 chain (Fig.~\ref{fig:quality}G), rankings in Figs.~\ref{fig:quality}B and C can be used to select the simulations giving the most realistic results in Fig.~\ref{fig:quality}A. (Figs.~\ref{fig:quality}E--F and~\ref{fig:POPC_POPE_dataSI} also show direct comparisons with the experimental data for the most relevant simulations; for reference, Fig.~\ref{fig:quality}D shows the overall highest-ranked simulation: POPC bilayer with OPLS3e parameters.) Based on these rankings, Lipid17 and Slipids simulations give the most realistic predictions for a POPC membrane, while simulations with CHARMM36 and GROMOS-CKP parameters predict overly packed bilayers (overestimated order in Fig.~\ref{fig:POPC_POPE_dataSI}). For POPE, on the other hand, GROMOS-CKP and Slipids give the most realistic results, while CHARMM36 and Lipid17 predict membranes that are too packed. In conclusion, the quality evaluation based on the NMRlipids Databank suggests that the Slipids parameters are the best currently available choice for simulations with PC and PE lipids, at least for applications where membrane packing is relevant.

@@

@@

\subsubsection{Extending the scope of MD simulations to new fields using NMRlipids Databank: Water diffusion anisotropy in membrane systems}
The anisotropy of water diffusion in directions parallel and perpendicular to membranes can be related to translocation of drugs through biological material, particularly in the skin \cite{hansen13,wen18,nitsche19,roberts21}, and to signal formation in diffusion tensor MRI imaging~\cite{topgaard20}. MD simulations are rarely used to analyze the anisotropic diffusion of water, since only a few membrane permeation events of water are typically observed in a single MD simulation trajectory~\cite{venable19,camilo2022}, thereby making the collection of a sufficient amount of data challenging. Here, we show that the API access to the data in NMRlipids Databank enables systematic analysis on how the anisotropic diffusion of water depends on membrane properties in multilamellar membrane systems, thereby extending the application of MD simulations to new fields.

@@

%The Discussion should be succinct and must not contain subheadings.

Sharing of biomolecular MD simulation and other data is becoming increasingly important in the age of big data and artificial intelligence~\cite{sever23,tiemann23}. 
%The main barriers for such solutions have been the required commitment for the long term support of hardware and software, and the lack of incentives for researchers to share the data. 
Besides the data itself, also programmatic access is a necessary requirement for data-driven and machine learning applications. This is particularly challenging when field-specific smart guidelines and community-consensus best-practices have not yet been defined, which is the case in biomolecular simulations~\cite{tiemann23}.
The NMRlipids Databank demonstrates how these issues can be solved by the overlay databank design, where the raw data are distributed to already publicly available decentralized locations, while the core of the databank is composed only of the metadata stored in a version-controlled git repository with an open-access license. On the other hand, the open-collaboration approach developed in the NMRlipids Project~\cite{botan15} creates incentives for sharing the data by offering authorship in published articles to the contributors. Advantages of such an approach are demonstrated here for membrane simulations, yet the concept can be applied to any other biomolecules, %such as disordered proteins and membrane--protein systems, 
as well as in any other field where similar barriers hinder the AI revolution, such as the assignment of NMR spectra~\cite{klukowski22}.

>>>>

\begin{abstract}
%TC:ignore
Tools based on artificial intelligence (AI) are currently revolutionising many fields, yet their applications are often limited by the lack of suitable training data in programmatically accessible format. Here we propose an effective solution to make data scattered in various locations and formats accessible for data-driven and machine learning applications using the overlay databank format. To demonstrate the practical relevance of such approach, we present the NMRlipids Databank---a community-driven, open-for-all database featuring programmatic access to quality-evaluated atom-resolution molecular dynamics simulations of cellular membranes. Cellular membrane lipid composition is implicated in diseases and controls major biological functions, but membranes are difficult to study experimentally due to their intrinsic disorder and complex phase behaviour. While MD simulations have been useful in understanding membrane systems, they require significant computational resources and often suffer from inaccuracies in model parameters. Here, we demonstrate how programmable interface for flexible implementation of data-driven and machine learning applications, and rapid access to simulation data through a graphical user interface, unlock possibilities beyond current MD simulation and experimental studies to understand cellular membranes. The proposed overlay databank concept can be further applied to other biomolecules, as well as in other fields where similar barriers hinder the AI revolution.

%Cellular membrane lipid composition is implicated in diseases and controls major biological functions, but membranes are difficult to study experimentally due to their intrinsic disorder and complex phase behaviour. Molecular dynamics (MD) simulations have been useful in understanding membrane systems, but they require significant computational resources and often suffer from inaccuracies in model parameters. Applications of data-driven and machine learning methods, currently revolutionising many fields, remain of limited use for membrane systems due to the lack of suitable training sets. Here we present the NMRlipids Databank---a community-driven, open-for-all database featuring programmatic access to quality-evaluated atom-resolution MD simulations of lipid bilayers. The NMRlipids Databank will benefit scientists in different disciplines by providing automatic ranking of simulations based on their quality against experiments, programmable interface for flexible implementation of data-driven and machine learning applications, and rapid access to simulation data through a graphical user interface. To demonstrate how it unlocks possibilities beyond current MD simulation studies, we analyzed the NMRlipids Databank to reveal how anisotropic diffusion of water and cholesterol flip-flop rates depend on membrane properties.
%TC:endignore
\end{abstract}\begin{document}

\flushbottom
%TC:ignore
\maketitle
%TC:endignore
% * <john.hammersley@gmail.com> 2015-02-09T12:07:31.197Z:
%
%  Click the title above to edit the author information and abstract
%
\thispagestyle{empty}

@@

Cellular membranes contain hundreds of different types of lipid molecules that regulate the membrane properties, morphology, and biological functions~\cite{vanmeer08,Lorent:2020a,Slatter:2016a}. Membrane lipid composition is implicated in diseases, such as cancer and neurodegenerative disorders, and therapeutics that affect membrane compositions are emerging~\cite{torres21}. However, biomembranes are often difficult to study experimentally, because they are complex mixtures of proteins and lipids in disordered fluid state with complicated phase behaviour at biological conditions. Molecular dynamics (MD) simulations can be used to model biomembranes in detail, but the computational cost for simulating all possible biological membrane compositions would be formidable. For those reasons, data-driven and machine-learning-based models that predict biomembrane properties will benefit wide range of fields covering academia and industry, from cell membrane biology to lipid nanoparticle formulations.

Here we present the NMRlipids Databank --- a community-driven, open-for-all database featuring programmatic access to atom-resolution MD simulations of lipid bilayers. To demonstrate its advantages over existing approaches, we build a machine learning model that predicts membrane properties from its lipid composition and show how information on rare phenomena that are beyond the scope of standard MD simulation investigations can be gleaned from the Databank. In addition, we demonstrate the immediate relevance of NMRlipids Databank in extending the scope of MD simulations to new fields: Using a data-driven approach, we are able to analyse how anisotropic diffusion of water depends on membrane properties; this benefits understanding in magnetic resonance imaging (MRI)~\cite{topgaard20} and pharmacokinetics~\cite{nitsche19}, where MD simulations have, until now, been rarely applied. Furthermore, the Databank performs automatic quality evaluation of membrane simulations, which facilitates the selection of best-performing models for each given application and accelerates the development of simulation parameters and methodology. Notably, the overlay-databank and open-collaboration approaches would facilitate the collecting of community-contributed data and providing programmatic access to them also in fields other than membrane simulations. For example, substantial biomolecular MD simulation data are already available but in scattered locations and formats~\cite{tiemann23}. The technical advances presented here will immediately benefit making these data programmatically accessible. Furthermore, the overlay databank configuration will benefit a broad range of fields where access to training data is a bottleneck for building AI-based tools.

@@

The overlay structure of the NMRlipids Databank, illustrated in Fig.~\ref{fig:overlay}A, is designed to enable efficient upcycling of MD simulations for data-driven and machine learning applications with minimal investment on new infrastructure. Raw simulation data in the {\it Data layer} can be stored in any publicly available location with long term stability and with permanent links to the data, such as digital object identifiers (DOIs), such as Zenodo (\href{https://zenodo.org}{zenodo.org}). The {\it Databank layer} (\href{https://github.com/NMRlipids/Databank}{github.com/NMRlipids/Databank}) is the core of the Databank containing all the relevant information about the simulations: links to the raw data, relevant metadata describing the systems, universal naming conventions for lipids and their atoms, quality evaluation of simulations against experimental data, and the computer programs to create the entries and to analyse the five basic properties extracted from all simulations (area per lipid, C--H bond order parameters, X-ray scattering form factors, membrane thickness, and equilibration times of principal components). Also the values for these five basic properties are stored in the {\it Databank layer}.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\linewidth]{Figures/overlay4.pdf}
    \caption{Overview of the NMRlipids Databank. 
    \textbf{A} Schematic presentation of the overlay structure used in the NMRlipids Databank. A more detailed structure of the \textit{Databank layer} is shown in Fig.~\ref{DatabankStructure} in the SI.
    \textbf{B} Distribution of lipids present in the trajectories of the Databank. `Others' lists lipids occurring in five or fewer simulations. 
    \textbf{C} Distribution of force fields in the simulations in the Databank. References for each force field are given in Table~\ref{tab:ForceFields}.
    \textbf{D} Flowchart for performing an analysis of properties through all MD simulations in the NMRlipids Databank using the API.
    \textbf{E} Flowchart for accessing results calculated from the NMRlipids Databank and stored to the \textit{Application layer}.
    \textbf{F} Currently available single lipid component and binary mixtures in the NMRlipids Databank. Colorbar shows the number of available simulations with the darkest green indicating three or more.  
    }
    \label{fig:overlay}
\end{figure}

@@

\subsection{NMRlipids Databank-API: programmatic access to the MD simulation data}\label{section:access}
The NMRlipids Databank-API provides programmatic access to all simulation data in the NMRlipids Databank through application programming interface (API). This enables wide range of novel data-driven applications---from construction of machine learning models that predict membrane properties, to automatic analysis of virtually any property across all simulations in the Databank. The flowchart in Fig.~\ref{fig:overlay}D illustrates the practical implementation of such an analysis. After cloning the databank repository to a local computer, raw data of each simulation can be accessed and analyzed with the help of functions delivered by the NMRlipids Databank-API. 
%by iterating through the README.yaml files in the {\it Databank layer} that contain links pointing to the locations of simulation trajectories and other relevant files. Simulations can then be automatically analyzed with the help of the README.yaml and the 
Analyses over simulations with different naming convetions can be automatically performed with the help of mapping files that associate the specific naming conventions in each simulation with the universal molecule and atom names used by the Databank. Finally, the analysis results can be stored using the same structure as in the {\it Databank layer}. Documentation of the NMRlipids Databank-API and further instructions are available at \href{https://nmrlipids.github.io/}{nmrlipids.github.io}, and template for new user-defined analyses at \href{https://github.com/NMRLipids/databank-template}{github.com/NMRLipids/databank-template}.
%Practical examples of codes that utilize the NMRlipids Databank-API, as well as a template for a new analysis, can be found from locations listed in Table~\ref{tab:codes}.

@@

\subsection{Applications of the NMRlipids Databank}
\subsubsection{Selecting simulation parameters using NMRlipids Databank: Best models for most abundant neutral membrane lipids}
Molecular dynamics (MD) simulations have been particularly useful in understanding membrane systems, although their accuracy has often been compromised by artefacts such as the quality of model parameters~\cite{antila22b,gupta22}. Presently, the accuracy of models is becoming increasingly important as researches are progressing from simulations of individual molecules to simulating whole organelles or even cells using interdisciplinary approaches~\cite{johnson15,thornburg22,gupta22}. Such systems exhibit intricate emergent behaviour making inaccuracies more difficult to detect, and accumulation of even modest errors may have a dramatic impact on the conclusions drawn. To minimize the detrimental consequences of artificial MD simulation results for their applications, the quality of lipid bilayer MD simulations has to be carefully assessed~\cite{antila22b}. This can be done, for example, against the C--H bond order parameters from NMR spectroscopy~\cite{bacle21,Wur23} and the form factors from X-ray scattering~\cite{ollila16}, although it requires comparisons between large number of simulations, which is laborious even with collaborative approaches~\cite{botan15,catte16,antila19,bacle21}.

Here we streamlined this process by defining quantitative quality measures 
for conformational ensembles of individual lipid molecules and membrane dimensions using C--H bond order parameters from NMR and X-ray scattering form factors~\cite{ollila16}. These measures enable automatic ranking of lipid bilayer simulations based on their quality against experiments. Qualities of order parameters were evaluated by first calculating the probabilities for each C--H bond order parameter to locate within experimental error, and then averaging the possibilities over different lipid segments ($P^{\mathrm{hg}}$, $P^{\mathrm{sn1}}$,$P^{\mathrm{sn2}}$, and $P^{\mathrm{total}}$). Qualities against X-ray scattering experiments ($FF_q$) were estimated as the difference in the experimental and simulated locations of the first form factor minimum. These measures are good proxies for membrane properties 
%can be used to evaluate membrane dimensions, as the form factor minima locations and acyl chain C--H bond order parameters 
because they correlate with the membrane lateral packing and thickness (Figs.~\ref{fig:quality}G,~\ref{fig:QualityCorrelationsSI}, and~\ref{fig:sizedependence}). Ergodicity of conformational sampling of lipids was estimated by calculating $\tau_\mathrm{rel}$, the convergence time of the slowest principal component divided by the simulation length.

Fig.~\ref{fig:quality} demonstrates how the automatic simulation-quality evaluation and the NMRlipids Databank-API enable rapid selection of the best models for membrane simulations. Fig.~\ref{fig:quality}A illustrates that predictions for the lateral packing of membranes composed of two most biologically-abundant neutral membrane lipids, POPC and POPE~\cite{vanmeer08},
%for area per lipid, one of the most important parameters used to characterize cellular membranes, 
diverge between different force fields. To find the most realistic parameters to simulate membranes with these lipids, we first ranked all simulations based on order parameter quality (Fig.~\ref{fig:top50simulations}), then picked force fields that occurr in Fig.~\ref{fig:quality}A (that is: force fields for both POPC and POPE in the Databank), and then ranked them according to the quality of the \textit{sn}-1 chain of POPC (Fig.~\ref{fig:quality}B) and of POPE (Fig.~\ref{fig:quality}C). Simulations with $\tau_\mathrm{rel}$ clearly above one (larger than 1.3) were discarded in this analysis. Because the average \textit{sn}-1 chain order parameter is a good proxy for the membrane packing (Fig.~\ref{fig:quality}G), rankings in Figs.~\ref{fig:quality}B and C can be used to select the simulations giving the most realistic results in Fig.~\ref{fig:quality}A. Based on this, Lipid17 and Slipids simulations are most realistic for a POPC membrane, while CHARMM36 and GROMOS-CKP simulations predict overly packed bilayers (overestimated order in Fig.~\ref{fig:POPC_POPE_dataSI}). For POPE, on the other hand, GROMOS-CKP and Slipids are most realistic, while CHARMM36 and Lipid17 predict too packed membranes. In conclusion, the quality evaluation based on the NMRlipids Databank suggests that the Slipids parameters are the best currently available choice for simulations with PC and PE lipids, at least for applications where membrane packing is relevant.
Also direct comparisons with the experimental data for the most relevant simulations are shown in Figs.~\ref{fig:quality}E--F and~\ref{fig:POPC_POPE_dataSI} a. Fig.~\ref{fig:quality}D shows the overall highest-ranked simulation, POPC bilayer with OPLS3e parameters, for the reference.

@@

\subsubsection{Using NMRlipids databank as a training set for machine learning applications: Predicting multi-component membrane properties}

To demonstrate the usage of the NMRlipids databank to construct machine learning (ML) models that predict membrane properties, we trained a model that predicts area per lipids and thicknesses of membranes with diverse compositions. First we used randomly selected 80\% of the databank simulations to choose the hyperparameters for, and optimize, a set of machine learning models with the goal of predicting the area per lipid from the membrane composition.

 After the parameter optimization, we tested predictions both against the remaining 20\% of the databank and against data from different ML models against area per lipids reported from simulations of membranes containing mixtures of POPC, POPE, POPS, PI, sphingomyelin lipids and cholesterol~\cite{shahane19,kumar21,oliveira22} not included in the databank. Linear regression and Ridge models gave the best correlation with the literature data in this test (predictions  from the linear regression model are shown in Fig.~\ref{fig:LinearRegressionTest}). The linear regression model was selected for further studies due to its simplicity.
\begin{figure}
    \centering
    \includegraphics[width=88mm]{Figures/LinearRegressionTest.pdf}
    \caption{Predictions of area per lipids of POPC, POPE, POPS, PI, sphingomyelin lipids and cholesterol from linear regression model against literature data from simulations (green~\cite{shahane19}, blue~\cite{kumar21}, and red~\cite{oliveira22}).}
    \label{fig:LinearRegressionTest}
\end{figure}

To demonstrate the usefulness of the constructed models for understanding multi-component membrane properties, we predicted later packing (ares per lipids) and membrane thicknesses of common biological membranes based on their lipid compositions reported in the literature, see table~\ref{tab:MLpredictions}. The model predicts substantial 50\% difference in area per lipid and 0.5\,nm in thickness between most densely (influenza virus) and loosely (mitochondria) packed membranes. Because these parameters correlate with many biologically relevant functions of membranes, such as permeation, cholesterol flip-flops (see next sections), and interactions with proteins~\cite{phillips09}, our results demonstrate that NMRlipids databank can be used to give valuable insights on biologically relevant properties of complex biological membranes. Most importantly, the delivered programmatic access to increasing amount of MD simulation data enables training of machine learning models that predict various membrane properties for all.

    \begin{table}[htb]
    \caption{Area per lipids (APL) and membrane thicknesses predicted by the linear regression model trained using the NMRlipids databank for membrane compositions corresponding different biological membranes. Systems are sorted in the order of increasing membrane packing. Compositions of different membranes are estimated based on references given in composition column.}
        \centering
        \begin{tabular}{cccc}
             Membrane&  Composition [lipid(\%-fraction)]&  APL[\AA$^2$] & Thickness [nm$^2$]\\
             \hline
             Mitochondria & POPC(37) : POPE(31) : PI(6) : CL(22) : 
             CHOL(4)~\cite{vanmeer08,escriba15,casares19} & 74 (66)$^*$ & 4.3 \\
             Bacterial & POPC(20) : POPE(35) : POPG(35) : CL(5) : CHOL(5)~\cite{chwastek20}  & 62 & 4.2 \\
             ER &  POPC(54) : POPE(20) : POPS(4) : PI(11) : SM(4) : CHOL(8)~\cite{vanmeer08,escriba15,casares19} & 56 & 4.5 \\
             Golgi & POPC(36) : POPE(21) : POPS(6) : PI(12) : SM(7) : CHOL(18)~\cite{vanmeer08,escriba15,casares19} & 51 & 4.7 \\
             Plasma & POPC(23) : POPE(11) : POPS(8) : PI(7) : SM(17) : CHOL(34)~\cite{vanmeer08,escriba15,casares19} & 45 & 4.9 \\
             Synaptic & POPC(28) : POPE(20) : POPS(9) : PI(2) : SM(4) : CHOL(37)~\cite{binotti21}  &  45 & 4.8 \\
             Influenza & POPC(5) : POPE(32) : POPS(15) : SM(9) : CHOL(40)~\cite{gerl12,ivanova15} & 42 & 4.8 \\
        \end{tabular}
        \label{tab:MLpredictions}

        $^*$ Value in parenthesis is the area per acyl chain taking into account that cardiolipin has four chains per molecule while other lipids have two.
    \end{table}

@@

\subsubsection{Extending the scope of MD simulations to new fields using NMRlipids Databank: Water diffusion anisotropy in membrane systems}
The anisotropic diffusion of water and hydrophilic molecules in directions parallel and perpendicular to membranes is important parameter in models describing the translocation of drugs through biological material, particularly in the skin \cite{hansen13,wen18,nitsche19,roberts21}. Water anisotropic diffusion plays a role also in the signal formation in diffusion tensor MRI imaging~\cite{topgaard20}. MD simulations are rarely used to analyze the anisotropic diffusion of water, since only a few membrane permeation events of water are typically observed in a single MD simulation trajectory~\cite{venable19,camilo2022}, thereby making the collection of a sufficient amount of data challenging. Here, we show that the API access to the data in NMRlipids Databank enables systematic analysis on how the anisotropic diffusion of water depends on membrane properties in multilamellar membrane systems, thereby extending the application of MD simulations to new fields.

@@

Sharing of biomolecular MD simulation and other data is becoming increasingly important in the age of big data and artificial intelligence~\cite{sever23,tiemann23}.  Besides the data itself, also programmatic access is a necessary requirement for data-driven and machine learning applications. This is particularly challenging when field-specific smart guidelines and community-consensus best-practices have not yet been defined, which is the case in biomolecular simulations~\cite{tiemann23}. The NMRlipids Databank demonstrates how these issues can be solved by the overlay databank design, where the raw data are distributed to already publicly available decentralized locations, while the core of the databank is composed only of the metadata stored in a version-controlled git repository with an open-access license. On the other hand, the open-collaboration approach developed in the NMRlipids Project~\cite{botan15} creates incentives for sharing the data by offering authorship in published articles to the contributors. Advantages of such an approach are demonstrated here for membrane simulations, yet the concept can be applied to any other biomolecules, as well as in any other field where similar barriers hinder the AI revolution, such as the assignment of NMR spectra~\cite{klukowski22}.
